Assessing the Firing of Ceramic Materials: a Seriation-Based Approach

### 2023-06-09 analysis/paper/paper.qmd
Small edit

---
title: "Assessing the Firing of Ceramic Materials: a Seriation-Based Approach"
author:
  - Nicolas Frerebeau:
      correspondence: "yes"
      email: nicolas.frerebeau@u-bordeaux-montaigne.fr
      orcid: 0000-0001-5759-4944
      affiliations:
        - name: "UMR 6034 Archéosciences Bordeaux"
          address: "Maison de l'Archéologie, Université Bordeaux Montaigne"
          city: Pessac cedex
          country: France
          postal-code: 33607
          url: https://www.archeosciences-bordeaux.fr
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  pdf: 
    documentclass: scrartcl
    include-in-header: 
      - text: |
          \usepackage{booktabs}
    classoption: 
      - french
      - 12pt
      - a4paper
      - oneside
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: [references.bib, packages.bib]
csl: "../templates/archeosciences.csl"
abstract: |
  The study of the firing of archaeological ceramics is thus a key factor in 
  interpreting ancient techniques and related social practices. This study 
  proposes a new method based on the correspondence analysis of powder X-ray 
  diffraction data (XRD), aiming to provide an alternative to the so-called
  Equivalent Firing Temperature framework. The properties of correspondence 
  analysis and its application for matrix seriation enable the description of 
  the firing of a ceramic assemblage. The application of correspondence 
  analysis to XRD data of Iron Age ceramics revealed the  significant potential 
  of this statistical method in the technological analysis of ancient pottery. 
  This allows to order of a large number of samples according to a firing 
  gradient and to track mineralogical transformations occurring during firing.
keywords: |
  X-ray diffraction; ceramic firing; correspondence analysis; matrix seriation
---

@@

```{r}
#| label: archaeological-data
## Read archaeological data
corpus <- here::here("analysis/data/raw_data/corpus.csv") |>
  read.table(sep = ",", dec = ".", header = TRUE, encoding = "UTF-8")
```

@@

```{r}
#| label: tbl-corpus
#| tbl-cap: "Ceramic and clay samples analyzed in this study. Phase 1 : 2nd century BC; Phase 2: first half of the 1st century BC; Phase 3: 50-30 BC. SU: stratigraphic unit. LOI: loss on ignition [see @frerebeau2020a; @frerebeau2015e; @frerebeau2015g]."
corpus[order(corpus$sample), !(colnames(corpus) %in% c("site_type", "decoration", "part"))] |> 
  knitr::kable(
    col.names = c("Sample", "Material", "Site", "SU", "Phase", "Type", "LOI (%)"),
    booktabs = TRUE,
    longtable = TRUE
  )
```

@@

## Ternary plot
xrf_cas <- here::here("analysis/data/derived_data/xrf_cas.csv") |> 
  read.table(header = TRUE, sep = ",", dec = ".") |> 
  subset(sample %in% corpus$sample)

par(mar = c(1, 1, 1, 1))
isopleuros::ternary_plot(
  xrf_cas[xrf_cas$site == "Mas de Moreno", c("CaO", "Al2O3", "SiO2")],
  col = "lightgrey", pch = 16, ylab = ""
)
isopleuros::ternary_points(
  xrf_cas[xrf_cas$site != "Mas de Moreno", c("CaO", "Al2O3", "SiO2")],
  pch = c(24, 25)[as.factor(xrf_cas$site[xrf_cas$site != "Mas de Moreno"])],
  bg = col_site[as.factor(xrf_cas$site[xrf_cas$site != "Mas de Moreno"])]
)
isopleuros::triangle_phase_ceramic(symbol = TRUE)
text(x = 0, y = 0.9, labels = "D", font = 2, cex = 1.5)
```

@@

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

>>>>

---
title: "Assessing the Firing of Ceramic Materials: a Seriation-Based Approach"
author:
  - Nicolas Frerebeau:
      correspondence: "yes"
      email: nicolas.frerebeau@u-bordeaux-montaigne.fr
      orcid: 0000-0001-5759-4944
      affiliations:
        - name: "UMR 6034 Archéosciences Bordeaux"
          address: "Maison de l'Archéologie, Université Bordeaux Montaigne"
          city: Pessac cedex
          country: France
          postal-code: 33607
          url: https://www.archeosciences-bordeaux.fr
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  pdf: 
    documentclass: scrartcl
    include-in-header: 
      - text: |
          \usepackage{booktabs}
    classoption: 
      - french
      - 12pt
      - a4paper
      - oneside
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    comment: "#>"
    fig.path: "../figures/"
    fig.dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: [references.bib, packages.bib]
csl: "../templates/archeosciences.csl"
abstract: |
  The study of the firing of archaeological ceramics is a key factor in 
  interpreting ancient techniques and related social practices. This study 
  proposes a new method based on the correspondence analysis of powder X-ray 
  diffraction data (XRD), aiming to provide an alternative to the so-called
  Equivalent Firing Temperature framework. The properties of correspondence 
  analysis and its application for matrix seriation enable the description of 
  the firing of a ceramic assemblage. The application of correspondence 
  analysis to XRD data of Iron Age ceramics revealed the  significant potential 
  of this statistical method in the technological analysis of ancient pottery. 
  This allows to order of a large number of samples according to a firing 
  gradient and to track mineralogical transformations occurring during firing.
  This method should let to more precise comparisons than those based on 
  estimating the equivalent firing temperature.
keywords: |
  X-ray diffraction; ceramic firing; correspondence analysis; matrix seriation
---

@@

```{r}
#| label: archaeological-data
## Read archaeological data
corpus <- here::here("analysis/data/raw_data/corpus.csv") |>
  read.table(sep = ",", dec = ".", header = TRUE, encoding = "UTF-8")

## Read XRF data
xrf_cas <- here::here("analysis/data/derived_data/xrf_cas.csv") |> 
  read.table(header = TRUE, sep = ",", dec = ".") |> 
  subset(sample %in% corpus$sample)
```

@@

```{r}
#| label: tbl-corpus
#| tbl-cap: "Ceramic and clay (unfired pottery sherds) samples analyzed in this study. Phase 1 : 2nd century BC; Phase 2: first half of the 1st century BC; Phase 3: 50-30 BC. LOI: loss on ignition [@frerebeau2020a; @frerebeau2015g; @frerebeau2015e]."
corpus |> 
  merge(y = xrf_cas[, -2], by = "sample") |> 
  (function(x) x[order(x$sample), ])() |> 
  (function(x) x[, !(colnames(x) %in% c("site_type", "stratigraphy", "artefact", "decoration", "part"))])() |> 
  (function(x) {
    write.table(x, file = here::here("analysis/tables/corpus.csv"), 
                col.names = TRUE, row.names = FALSE, sep = ",", dec = ".")
    x
  })() |> 
  knitr::kable(
    col.names = c("Sample", "Material", "Site", "Phase", "LOI (%)", "CaO (%)", "SiO2 (%)", "Al2O3 (%)"),
    booktabs = TRUE,
    longtable = TRUE
  )
```

@@

## Ternary plot
par(mar = c(1, 1, 1, 1))
isopleuros::ternary_plot(
  xrf_cas[xrf_cas$site == "Mas de Moreno", c("CaO", "Al2O3", "SiO2")],
  col = "lightgrey", pch = 16, ylab = ""
)
isopleuros::ternary_points(
  xrf_cas[xrf_cas$site != "Mas de Moreno", c("CaO", "Al2O3", "SiO2")],
  pch = c(24, 25)[as.factor(xrf_cas$site[xrf_cas$site != "Mas de Moreno"])],
  bg = col_site[as.factor(xrf_cas$site[xrf_cas$site != "Mas de Moreno"])]
)
isopleuros::triangle_phase_ceramic(symbol = TRUE)
text(x = 0, y = 0.9, labels = "D", font = 2, cex = 1.5)
```

@@
